Model matters with restricted Boltzmann machines

### 2017-04-17 presentations/coms474/slides.Rmd
guest lecture

---
title: "Model matters with restricted Boltzmann machines"
shorttitle: "Model matters with RBMs"
author: Andee Kaplan
shortname: Kaplan, et al.
institute: |
    | Iowa State University
    | ajkaplan@iastate.edu
shortinstitute: ajkaplan@iastate.edu
date: |
  | February 14, 2017
  |
  | Slides available at <http://bit.ly/kaplan-msu>
  | 
  | \footnotesize Joint work with D. Nordman and S. Vardeman
shortdate: "February 14, 2017"
output: 
  beamer_presentation:
    template: ../resources/latex/beamer.tex
    includes:
      in_header: ../resources/latex/front-matter.tex
theme: CambridgeUS
bibliography: ../resources/latex/refs.bib
fig_caption: true
nocite: |
    @salakhutdinov2009deep, @kaiser2007statistical
---

@@

# RBM quantities to compare

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)} \\
~\\
\text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)} 
\end{align*}

# Data coding to mitigate degeneracy

\begin{figure}
\includegraphics[width=.55\linewidth]{../resources/images/toyhull_top.png}
\includegraphics[width=.35\linewidth]{../resources/images/toyhull_top_2.png}
\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
\end{figure}

The convex hull of $\mathcal{T} \subset \mathcal{C}^3$ does not fill the unit cube $[0,1]^3$ (left), but does better with $[-1,1]^3$ (right).

@@

- For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space.
- At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!

```{r encoding-volume, echo = FALSE, fig.height=3, fig.cap="\\label{fig:volume_plot}Relationship between volume of the convex hull of possible values of the RBM sufficient statistics and the cube containing it for different size models."}
test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
  rename(H = X1.4, V = X1.4.1) %>%
  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
  mutate(n_param = H*V + H + V) %>%
  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently

bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))

plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
  mutate(frac_vol = V1/(1^n_param)) %>%
  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
               mutate(frac_vol = V1/(2^n_param)),
             by="n_param") %>%
  rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
  gather(vars, value, -n_param) %>%
  separate(vars, c("type", "encoding"), "\\.") %>%
  spread(type, value) %>%
  ggplot() +
  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
  ylab("Fraction of unrestricted volume") +
  xlab("Number of parameters") +
  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)")) +           
  theme(
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA)) -> p

p
```

@@

# Hyperparameters

@@

# Future work

1. Generalization of instability results for other network models [ongoing, see @kaplan2016note]
2. Image classification
    - Ensemble methods (super learners) using AdaBoost [@freund1995desicion] 
    - Decision theoretic based approach to approximating the likelihood ratio test for classification
3. Markov chain Monte Carlo methods for data with Markovian dependence
    - Spatial data
    - Network data

@@

* Slides -- <http://bit.ly/kaplan-msu>

@@

>>>>

---
title: "Model matters with restricted Boltzmann machines"
shorttitle: "Model matters with RBMs"
author: Andee Kaplan
shortname: Kaplan, et al.
institute: |
    | Iowa State University
    | ajkaplan@iastate.edu
shortinstitute: ajkaplan@iastate.edu
date: |
  | April 17, 2017
  |
  | Slides available at <http://bit.ly/kaplan-coms474>
  | 
  | \footnotesize Joint work with D. Nordman and S. Vardeman
shortdate: "April 17, 2017"
output: 
  beamer_presentation:
    template: ../resources/latex/beamer.tex
    includes:
      in_header: ../resources/latex/front-matter.tex
theme: CambridgeUS
bibliography: ../resources/latex/refs.bib
fig_caption: true
nocite: |
    @salakhutdinov2009deep, @kaiser2007statistical
---

@@

@@

- For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space.
\vspace{.6in}
- At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!

@@

@@

@@

* Slides -- <http://bit.ly/kaplan-coms474>

@@

# Appendix: Hyperparameters
